Signatures and Induction Principles for Higher Inductive-Inductive Types

### 2019-01-28 paper.tex
RECOLOR PARENS TO RED

Again, let us consider the circle example:
\begin{alignat*}{5}
  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,(\hat{\tt,\,S^1,\,b,\,loop})\\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1D}}\in \hat{S^1}\rah\Type_i)\timesh(\hat{b^D}\in \hat{S^{1D}\,b})\timesh(\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D})
\end{alignat*}
The inputs of $\blank^\D$ here are the signature for the circle (the context in
black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
It returns a family over the type $\hat{S^1}$, an element of this family
$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.

@@

TODO: revise when we have homomorphism section, in order to remove redundant
explanations. Also revise language.

@@

\infer{\hat{\Gamma}\semicol\Delta^\S \in (\hat{\gamma}\in\Delta^\A)\rah \Delta^\D\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}

@@

Types are interpreted as dependent binary relations which additionally
depend on ($\hat{\gamma}\commah\,\hat{\gamma^D}\commah\,\hat{\gamma^S}$) interpretations of the

@@

\infer{\hat{\Gamma}\vdashh A^\S \in (\hat{\gamma}\in\Delta^\A)(\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma})(\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D})(\hat{x}\in A^\A\,\hat{\gamma})\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}

@@

\infer{\hat{\Gamma}\vdashh t^\S \in (\hat{\gamma}\in\Delta^\A)(\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma})(\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D})\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,(t^\A\,\hat{\gamma})\,(t^\D\,\hat{\gamma}\,\hat{\gamma^D})}{\hat{\Gamma}\semicol\Delta \vdash t : A}

@@

%% \begingroup \allowdisplaybreaks

@@

  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv (\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2})\timesh A^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} \\

@@

  & \U^\S\,\hat{\gamma^3}\,\hat{A}\,\hat{A^D} && :\equiv (\hat{x}\in \hat{A})\rah  \hat{A^D}\,\hat{x} \\

@@

  & ((x:a)\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv (\hat{x}\in a^\A\,\hat{\gamma})\rah  B^\S\,(\hat{\gamma,x})\,(\hat{\gamma^D,}\,a^\S\,\hat{\gamma^3}\,\hat{x})\,(\hat{\gamma^S,}\,\hat{\refl}) \\
  & && \hspace{8.4em} (\hat{f}\,\hat{x})\,\big(\hat{f^D}\,\hat{x}\,(a^\S\,\hat{\gamma^3}\,\hat{x})\big) \\
  & (t\,u)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,(t^\S\,\hat{\gamma^3}\,( u^\A\,\hat{\gamma}))\,(u^\S\,\hat{\gamma^3}) \\
  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv (\hat{x}\in \hat{A})\rah B^\S\,\hat{\gamma^3}\,(\hat{f}\,\hat{x})\,(\hat{f^D}\,\hat{x}) \\

@@

  & (t=_a u)^\S\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,(t^\S\,\hat{\gamma^3})\big(\TR\,(u^\S\,\hat{\gamma^3})\,(\apd\,(a^\S\,\hat{\gamma^3})\,\hat{e})\big) \\
  & (\refl_t)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,(t^\S\,\hat{\gamma^3}) \\

@@

  & && \hspace{-10em}\hat{\J}\,\bigg(\hat{\J}\,\Big(\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S})\,(t^\S\,\hat{\gamma^3})\,\\
  & && \hspace{-9em}(\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3})\,(pr^\D\,\hat{\gamma^2})\,(pr^\S\,\hat{\gamma^3})\big)\,eq^\A\,\hat{\gamma}\Big)\,u^\S\,\hat{\gamma^3}\bigg)\,(eq^\S\,\hat{\gamma^3}) \\

@@

  & && \hspace{-10em}\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl})\,(t^\S\,\hat{\gamma^3})\,(\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3})\big)\,(pr^\S\,\hat{\gamma^3})\\
  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^3\,f\,t} && :\equiv b^\S\,\hat{\gamma^3}\,(\hat{f}\,\hat{t}) \\
  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv \ap\,(\hat{\lambda} \hat{f.}\hat{f}\,\hat{u})\,(t^\S\,\hat{\gamma^3})

@@

The $\U^\S$ rule yields the type of the eliminator for a type
formation rule. In the natural numbers example above, the non-indexed
$Nat : \U$ sort is interpreted as $\hat{n^S} \in (\hat{x}\in
\hat{n})\rah \hat{n^D}\,\hat{x}$. For indexed sorts, the indices are
first processed by the $\blank^\S$ cases for inductive and non-inductive
parameters, until the ultimate $\U$ return type is reached. Hence, the
eliminator for a sort is always a function.

@@

The circle example is a bit more interesting here:
\begin{alignat*}{5}
  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,(\hat{\tt,\,S^1,\,b,\,loop})\,(\hat{\tt,\,S^{1M},\,b^D,\,loop^D}) \\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1E}}\in (\hat{x}\in\hat{S^1})\hat{\,\ra\,}\hat{S^{1M}\,x})\timesh(\hat{b^S}\in \hat{S^{1E}\,\,b = b^D})\\
  & && \hspace{0.75em}\timesh(\hat{loop^S\,\,\in\,\,} \TR\,_{(\hat{\lambda x. \TR\,_{S^{1M}}\,loop\,x\,=\,b^D})}\,\hat{b^S}\,(\TR\,_{(\hat{\lambda} \hat{x}. \TR\,_{\hat{S^{1M}}}\,\hat{loop}\,(\hat{S^{1E}\,b})\,\eqh\,\hat{x})}\,\hat{b^S}\,(\apd\,\hat{S^{1E}\,loop}))\,\\
  & && \hspace{5.5em} \hat{=\,loop^D})
\end{alignat*}
In homotopy type theory, the $\beta$-rule for $loop$ is usually just
$\apd\,\hat{S^{1E}\,loop}\,\hat{\,=\,loop^D}$, but here all
$\beta$-rules are propositional, so we need to transport with
$\hat{b^S}$ to make the equation well-typed. When computing the type
of $\hat{loop^S}$, we start with $(\underline{b
  =b})^S\,\,\hat{\gamma^3}\,\,\hat{loop}\,\,\hat{loop^D}$. Next, this
evaluates to $(b=b)^\S\,\hat{\gamma^3}\,\hat{loop}\eqh\hat{loop^D}$,
and then we unfold the left hand side to get the doubly-transported
expression in the result.

@@

For another example for the translations, we consider indexed W-types
which can describe a large class of inductive definitions
\cite{morris09indexed}. Suppose we are in the target context $\hat{I}
\in \Type_{\hat{0}}\commah\hat{S} \in \Type_{\hat{0}}\commah\hat{P} \in \hat{S} \rah
\Type_{\hat{0}}\commah\hat{out} \in \hat{S} \rah \hat{I}\commah\hat{in} \in
(\hat{s} \in \hat{S})\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
code for the corresponding indexed W-type is the following:
\[
W :\equiv (\boldsymbol{\cdot},\,\,\,w:(\hat{i} \in \hat{I})\ra\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P\,s})\ra w\,(\hat{in\,s\,p}))\ra \underline{w\,(\hat{out\,s})})
\]
We pick a universe level $j$ for elimination. The interpretations of $W$ are the following, omitting leading $\hat{\top}$ components:
\begin{alignat*}{5}
  & \rlap{$W^\A \equiv (\hat{w}\in \hat{I}\rah\Type_{\hat{0}})\timesh((\hat{s}\in \hat{S})\rah((\hat{p}\in \hat{P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\rah \hat{w}\,(\hat{out\,s}))$}\\
  & W^\D\,(\hat{w, sup}) && \equiv\, && (\hat{w^D} \in (\hat{i}\in \hat{I})\rah \hat{w\,i}\rah \Type_j)\\
  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
  & && && \rah ((\hat{p\in P\,s})\rah \hat{w^D}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\rah \hat{w^D}\,(\hat{out\,s})\,(\hat{sup\,s\,f})\big) \\
  & \rlap{$ W^\S\,(\hat{w, sup})\,(\hat{w^D,sup^D}) \equiv $} \\
  & && && (\hat{w^S} \in (\hat{i \in I})(\hat{x \in w\,i})\rah \hat{w^D\,i\,x})\\
  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
  & && && \rah \hat{w^S}\,(\hat{out\,s})\,(\hat{sup\,\,s\,f}) \eqh \hat{sup^D\,s\,f}\,(\hat{\lambda} \hat{p.}\, \hat{w^S}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\big)
\end{alignat*}

@@

The two-dimensional sphere is given by the following context in the
theory of codes.
\[
\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
\]
The sphere-algebras are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\C \equiv \hat{\top}\timesh(\hat{S^2}\in\Type_{\hat{0}})\timesh(\hat{b}\in \hat{S^2})\timesh(\hat{surf\in \hat{\refl_{\hat{b}}}=_{(\hat{b}=_{\hat{S^2}}\hat{b})}\hat{\refl_{\hat{b}}}})
\end{alignat*}
Given a sphere-algebra and fixing a universe level $i$, the motives
and methods are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\M\,\,(\hat{\tt,\,S^2,\,b,\,surf})\\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2M}}\in \hat{S^2}\rah\Type_i) \\
  & && \hspace{0.75em} \timesh(\hat{b^M}\in \hat{S^{2M}\,b}) \\
  & && \hspace{0.75em} \timesh(\hat{surf^M\in \TR_{(\TR_{S^{2M}}\,\blank\,b^M=b^M)}\,surf\,\refl_{b^M} = \refl_{b^M}})
\end{alignat*}
Given a sphere-algebra and the motives and methods for this algebra,
the types of the elimination principles are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\E\,\,(\hat{\tt,\,S^2,\,b,\,surf})\,(\hat{\tt,\,S^{2M},\,b^M,\,surf^M}) \\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2E}}\in (\hat{x}\in\hat{S^2})\hat{\,\ra\,}\hat{S^{2M}\,x}) \\
  & && \hspace{0.75em} \timesh(\hat{b^E}\in \hat{S^{2E}\,\,b = b^M})\\
  & && \hspace{0.75em} \timesh\bigg(\hat{surf^E\,\,\in\,\,} \TR\,\Big(\hat{\J\,\refl\,b^E}\Big)\,\Big(\TR\,\big(\hat{\J\,\refl\,b^E}\big)\,\big(\hat{\apd}\,(\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,(\TR\,\hat{b^E}\,(\hat{\apd}\,\hat{S^{2E}}\,\hat{x})))\,\hat{surf}\big)\Big) \\
  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\bigg)
\end{alignat*}
Note that if $\hat{b^E}$ is a definitional equality (that is, we have
$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
this case the type of $\hat{surf^E}$ becomes the expected
$\hat{\apd}\,(\hat{\apd}\,\hat{S^{2E}})\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.

>>>>

Again, let us consider the circle example:
\begin{alignat*}{5}
  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_i\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
\end{alignat*}
The inputs of $\blank^\D$ here are the signature for the circle (the context in
black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
It returns a family over the type $\hat{S^1}$, an element of this family
$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.

@@

TODO: REWRITE to remove redundacy from homomorphism section.

@@

\infer{\hat{\Gamma}\semicol\Delta^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}

@@

Types are interpreted as dependent binary relations which additionally depend on
$\hat{\gamma}$, $\hat{\gamma^D}$, $\hat{\gamma^S}$ interpretations of the

@@

\infer{\hat{\Gamma}\vdashh A^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\hat{(}\hat{x}\in A^\A\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}

@@

\infer{\hat{\Gamma}\vdashh t^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\D\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}}{\hat{\Gamma}\semicol\Delta \vdash t : A}

@@

\begingroup
%% \allowdisplaybreaks

@@

  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} \\

@@

  & \U^\S\,\hat{\gamma^3}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\

@@

  & ((x:a)\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^3}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^3}\,\hat{x}\hat{)}\hat{\big)} \\
  & (t\,u)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^3}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^3}\hat{)} \\
  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^3}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\

@@

  & (t=_a u)^\S\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^3}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
  & (\refl_t)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)} \\

@@

  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\,\\
  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^3}\hat{)}\hat{\big)}\,eq^\A\,\hat{\gamma}\hat{\Big)}\,u^\S\,\hat{\gamma^3}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^3}\hat{)} \\

@@

  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^3}\hat{)}\\
  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^3\,f\,t} && :\equiv b^\S\,\hat{\gamma^3}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}

@@

The $\U^\S$ rule yields the type of the eliminator for a type
formation rule. In the natural numbers example above, the non-indexed
$Nat : \U$ sort is interpreted as $\hat{\hat{n^S} \in (\hat{x}\in
\hat{n})\rah \hat{n^D}\,\hat{x}}$. For indexed sorts, the indices are
first processed by the $\blank^\S$ cases for inductive and non-inductive
parameters, until the ultimate $\U$ return type is reached. Hence, the
eliminator for a sort is always a function.

@@

The circle example is a bit more interesting here:
\begin{alignat*}{5}
  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\,\hat{(}\hat{\tt,\,S^{1M},\,b^D,\,loop^D}\hat{)} \\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1E}}\in \hat{(}\hat{x}\in\hat{S^1}\hat{)}\hat{\,\ra\,}\hat{S^{1M}\,x}\hat{)}\timesh\hat{(}\hat{b^S}\in \hat{S^{1E}\,\,b = b^D}\hat{)}\\
  & && \hspace{0.75em}\timesh\hat{(}\hat{loop^S\,\,\in\,\,} \TR\,_{\hat{(}\hat{\lambda x. \TR\,_{S^{1M}}\,loop\,x\,=\,b^D}\hat{)}}\,\hat{b^S}\,\hat{(}\TR\,_{\hat{(}\hat{\lambda} \hat{x}. \TR\,_{\hat{S^{1M}}}\,\hat{loop}\,\hat{(}\hat{S^{1E}\,b}\hat{)}\,\eqh\,\hat{x}\hat{)}}\,\hat{b^S}\,\hat{(}\apd\,\hat{S^{1E}\,loop}\hat{)}\hat{)}\,\\
  & && \hspace{5.5em} \hat{=\,loop^D}\hat{)}
\end{alignat*}
In homotopy type theory, the $\beta$-rule for $loop$ is usually just
$\apd\,\hat{S^{1E}\,loop}\,\hat{\,=\,loop^D}$, but here all
$\beta$-rules are propositional, so we need to transport with
$\hat{b^S}$ to make the equation well-typed. When computing the type
of $\hat{loop^S}$, we start with $(\underline{b
  =b})^S\,\,\hat{\gamma^3}\,\,\hat{loop}\,\,\hat{loop^D}$. Next, this
evaluates to $(b=b)^\S\,\hat{\gamma^3}\,\hat{loop}\eqh\hat{loop^D}$,
and then we unfold the left hand side to get the doubly-transported
expression in the result.

@@

For another example for the translations, we consider indexed W-types
which can describe a large class of inductive definitions
\cite{morris09indexed}. Suppose we are in the external context $\hat{I}
\in \Type_{\hat{0}}\commah\hat{S} \in \Type_{\hat{0}}\commah\hat{P} \in \hat{S} \rah
\Type_{\hat{0}}\commah\hat{out} \in \hat{S} \rah \hat{I}\commah\hat{in} \in
\hat{(}\hat{s} \in \hat{S}\hat{)}\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
code for the corresponding indexed W-type is the following:
\[
W :\equiv (\boldsymbol{\cdot},\,\,\,w:(\hat{i} \in \hat{I})\ra\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P\,s})\ra w\,(\hat{in\,s\,p}))\ra \underline{w\,(\hat{out\,s})})
\]
We pick a universe level $j$ for elimination. The interpretations of $W$ are the following, omitting leading $\hat{\top}$ components:
\begin{alignat*}{5}
  & \rlap{$W^\A \equiv \hat{(}\hat{w}\in \hat{I}\rah\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{(}\hat{s}\in \hat{S}\hat{)}\rah\hat{(}\hat{(}\hat{p}\in \hat{P\,s}\hat{)}\rah \hat{w}\,\hat{(}\hat{in\,s\,p}\hat{)}\hat{)}\rah \hat{w}\,\hat{(}\hat{out\,s}\hat{)}\hat{)}$}\\
  & W^\D\,\hat{(}\hat{w, sup}\hat{)} && \equiv\, && \hat{(}\hat{w^D} \in \hat{(}\hat{i}\in \hat{I}\hat{)}\rah \hat{w\,i}\rah \Type_{\hat{j}}\hat{)}\\
  & && \hspace{0.23em}\timesh && \hat{\big(}\hat{(}\hat{s\in S}\hat{)}\hat{(}\hat{f}\in \hat{(}\hat{p\in P\,s}\hat{)}\rah \hat{w}\,\hat{(}\hat{in\,s\,p}\hat{)}\hat{)}\\
  & && && \rah \hat{(}\hat{(}\hat{p\in P\,s}\hat{)}\rah \hat{w^D}\,\hat{(}\hat{in\,s\,p}\hat{)}\,\hat{(}\hat{f\,p}\hat{)}\hat{)}\rah \hat{w^D}\,\hat{(}\hat{out\,s}\hat{)}\,\hat{(}\hat{sup\,s\,f}\hat{)}\hat{\big)} \\
  & \rlap{$ W^\S\,\hat{(}\hat{w, sup}\hat{)}\,\hat{(}\hat{w^D,sup^D}\hat{)} \equiv $} \\
  & && && \hat{(}\hat{w^S} \in \hat{(}\hat{i \in I}\hat{)}\hat{(}\hat{x \in w\,i}\hat{)}\rah \hat{w^D\,i\,x}\hat{)}\\
  & && \hspace{0.23em}\timesh && \hat{\big(}\hat{(}\hat{s\in S}\hat{)}\hat{(}\hat{f}\in \hat{(}\hat{p\in P\,s}\hat{)}\rah \hat{w}\,\hat{(}\hat{in\,s\,p}\hat{)}\hat{)}\\
  & && && \rah \hat{w^S}\,\hat{(}\hat{out\,s}\hat{)}\,\hat{(}\hat{sup\,\,s\,f}\hat{)} \eqh \hat{sup^D\,s\,f}\,\hat{(}\hat{\lambda} \hat{p.}\, \hat{w^S}\,\hat{(}\hat{in\,s\,p}\hat{)}\,\hat{(}\hat{f\,p}\hat{)}\hat{)}\hat{\big)}
\end{alignat*}

@@

The two-dimensional sphere is given by the following context in the
theory of codes.
\[
\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
\]
The sphere-algebras are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\C \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
\end{alignat*}
Given a sphere-algebra and fixing a universe level $i$, the motives
and methods are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\M\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2M}}\in \hat{S^2}\rah\Type_i\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{b^M}\in \hat{S^{2M}\,b}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^M\in \TR_{\hat{(}\TR_{S^{2M}}\,\blank\,b^M=b^M\hat{)}}\,surf\,\refl_{b^M} = \refl_{b^M}}\hat{)}
\end{alignat*}
Given a sphere-algebra and a displayed algebra over it, we get the type of sections:
\begin{alignat*}{5}
  & && \Gamma^\E\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2M},\,b^M,\,surf^M}\hat{)} \\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2E}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2M}\,x}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{b^E}\in \hat{S^{2E}\,\,b = b^M}\hat{)}\\
  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^E\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^E}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^E}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,\hat{(}\TR\,\hat{b^E}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\hat{\bigg)}
\end{alignat*}
Note that if $\hat{b^E}$ is a definitional equality (that is, we have
$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
this case the type of $\hat{surf^E}$ becomes the expected
$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
